Curve counting in genus one: elliptic singularities {\it \&} relative geometry

### 2019-06-28 EllipticSingularitiesSplitting.tex
implemented Dhruv's changes to introduction

While the higher genus situation is much more difficult, in the last several years, substantial theory has been developed towards a conceptual understanding of the genus one case. This began over a decade ago with the pioneering work of Li, Vakil, and Zinger~\cite{redgone,VZ,LZ,lz2,zingerstvsred,zingred}, and was recently reinvigorated involving ideas from logarithmic geometry and singularity theory~\cite{BCM18,CM18,HL,RSPW,RSPW2,SMY1,VISC}. The result is a \textbf{reduced} Gromov--Witten theory, which removes certain degenerate contributions from the ordinary theory. While the Atiyah--Bott localization for the reduced invariants was used to great effect by Zinger in his proof of the BCOV conjectures, two major components of the standard theory -- relative invariants and the degeneration formula -- have not been developed in this setting. Our results in this paper provide these as new {structural} tools in reduced genus one Gromov--Witten theory\footnote{Further motivation to consider the genus one case specifically comes simply from the fact that for Calabi--Yau geometries in dimension larger than $3$, Gromov--Witten theory vanishes in genera greater than one.}.

\subsection{Results} Our first contribution is a definition and construction of relative reduced Gromov--Witten theory of projective space relative to a hyperplane. The geometric content is the following, see \S \ref{section construction} for a precise formulation.

@@

\begin{customthm}{A}\label{thm: desingularization}
There exists a logarithmically smooth and proper Deligne--Mumford stack $\VZ_{1,\alpha}^\circ(\mathbb P^N|H,d)$ having expected dimension, and containing $\mathcal M_{1,\alpha}^\circ(\mathbb P^N|H,d)$  as a dense open subset. A point in this moduli space parameterizes maps
\[
C\to \mathbb P^N[s]\to \mathbb P^N,
\]
to an $s$-fold expansion of $\mathbb P^N$ along $H$ such that the map to the expansion and composite map $C\to \mathbb P^N$ each factor nontrivially through an elliptic singularity. 
\end{customthm}

@@

\begin{customthm}{B}\label{thm: recursion}
Given a smooth pair $(X,Y)$ with $Y$ very ample, there is an explicit recursive algorithm to calculate
\begin{enumerate}
\item the (restricted) reduced genus one Gromov--Witten invariants of $Y$;
\item the reduced genus one relative Gromov--Witten invariants of $(X,Y)$; 
\item the (restricted) reduced genus one rubber invariants of $\mathbb{P}=\PP_Y(\operatorname{N}_{Y|X} \oplus \OO_Y)$
\end{enumerate}
from the full genus zero and reduced genus one invariants of the ambient variety $X$.
\end{customthm}

We pursue a strategy laid out by Vakil and Gathmann, building on work of Caporaso and Harris~\cite{CH98,Ga,Vre}. The first step is to express the locus of maps where the contact order degenerates in terms of Chern classes of natural vector bundles. The second step is to describe that locus in terms of fibre products of moduli spaces with smaller invariants. Novelties are introduced in both steps. We achieve the first step using techniques that are entirely within tropical and logarithmic geometry, drastically simplifying Gathmann's calculation. For the second step, we contend with the interaction of the relative splitting formula and the geometry of elliptic singularities, which is entirely new (\S \ref{subsection C0 splitting}). We express the factorisation property in terms of tautological classes and combinatorial manipulations, leading to Theorem~\ref{thm: recursion}.

\subsection{Broader contributions} Beyond genus one, we pass through a number of seemingly useful general techniques. First, we explain how to describe degenerate moduli spaces as fibre products. This is substantially more delicate than the standard situation because of the need to work with elliptic singularities and aligned logarithmic structures. This interaction of the degeneration formula with the geometry of curve singularities is a new phenomenon. The results of~\cite{RSPW} suggest the existence of a reduced higher genus Gromov--Witten theory formed by replacing contracted elliptic components with singularities\footnote{We have been informed by Jonathan Wise that his Ph.D. candidate Sebastian Bozlee has constructed these higher genus reduced Gromov--Witten invariants. A manuscript is in preparation.}. The discussion here is likely to carry over to this setting.

Second, a key step in our recursion is a description of the locus of maps with higher than prescribed tangency, which was identified by Gathmann. We realize the locus as the vanishing locus of a section of a line bundle that comes from tropical geometry — from a piecewise linear function on the fan. The systematic understanding of logarithmic line bundles arising in this fashion is likely to play an important r\^ole in logarithmic enumerative geometry.

\tred{We rephrase Gathmann's idea of increasing the tangency requirement within the framework of logarithmic stable maps (where maximal tangency is always assumed) by introducing \emph{fictitious markings}, and identifying the space of maps with higher contact order with a boundary stratum inside that of lower contact order and more (fictitious) markings.}

@@

Finally, we note that our results require us to encounter several variants of the space of relative/logarithmic stable maps, with both fixed and expanded target. The conceptual features of one space are often computational bugs, and vice versa, and we believe that the techniques developed in the genus one case here will be used repeatedly in logarithmic Gromov--Witten theory calculations. In particular, we frequently use Kato's perspective on logarithmic blowups as subfunctors \cite{KatoExactness} that was applied in~\cite{RSPW,RSPW2}.

\subsection*{Acknowledgements} We have benefited from fruitful conversations with friends and colleagues, including Pierrick Bousseau, Francesca Carocci, Cristina Manolache, Davesh Maulik and Jonathan Wise. Important aspects of this work were completed while L.B. and N.N. were visiting MIT in Spring 2018 and the University of Cambridge in Winter 2019, and we thank these institutions for their hospitality. L.B. and N.N. would like to thank Imperial College London, the University of Glasgow, the LSGNT, the Royal Society and 3CinG for support. L.B. would in addition like to thank the MSRI program ''Enumerative Geometry Beyond Numbers'' and MPIM-Bonn for support.

\newpage

@@

\noindent
\textbf{Step 3}.\marginpar{doesn' picking $\delta$ itself imply a subdivision?} Given a tropical map $F: \plC\to \mathbb R_{\geq 0}$ parameterized by $\Sigma^{\mathrm{Kim}}$, there is a largest radius $\delta$ (possibly equal to $0$) such that every vertex strictly contained in the circle of radius $\delta$ around the core has degree-marking $d=0$. Let $
\Sigma^{\mathrm{cen}}$ be the subdivision obtained by requiring that the vertices contained within the circle of radius $\delta$ around the core of $\plC$ are totally ordered. This involves introducing cones along which certain $\Z$-linear combinations of edge lengths are identified.\\

>>>>

While the higher genus situation is much more difficult, in the last several years, substantial theory has been developed towards a conceptual understanding of the genus one case. This began over a decade ago with the pioneering work of Li, Vakil, and Zinger~\cite{redgone,VZ,LZ,lz2,zingerstvsred,zingred}, and was recently reinvigorated involving ideas from singularity theory~\cite{BCM18,CM18,HL,SMY1,VISC} and logarithmic geometry~\cite{RSPW,RSPW2}. The result is a \textbf{reduced} Gromov--Witten theory, which removes certain degenerate contributions from the ordinary theory. While the Atiyah--Bott localization for the reduced invariants was used to great effect by Zinger in his proof of the BCOV conjectures, two major components of the standard theory -- relative invariants and the degeneration formula -- have not been developed in this setting. Our results in this paper provide these as new {structural} tools in reduced genus one Gromov--Witten theory\footnote{Further motivation to consider the genus one case specifically comes simply from the fact that for Calabi--Yau geometries in dimension larger than $3$, Gromov--Witten theory vanishes in genera greater than one.}.

\subsection{Results} Our first contribution is a definition and construction of reduced Gromov--Witten theory of projective space relative to a hyperplane. The geometric content is the following, see \S \ref{section construction} for a precise formulation.

@@

\begin{customthm}{A}\label{thm: desingularization}
There exists a logarithmically smooth and proper Deligne--Mumford stack $\VZ_{1,\alpha}(\mathbb P^N|H,d)$ having expected dimension, and containing $\mathcal M_{1,\alpha}^\circ(\mathbb P^N|H,d)$  as a dense open subset. A point in this moduli space parameterizes maps
\[
C\to \mathbb P^N[s]\to \mathbb P^N,
\]
to an $s$-fold expansion of $\mathbb P^N$ along $H$ such that the map to the expansion and composite map $C\to \mathbb P^N$ each factor nontrivially through an elliptic singularity. 
\end{customthm}

@@

\begin{customthm}{B}\label{thm: recursion}
Given a smooth pair $(X,Y)$ with $Y$ very ample, there is an explicit recursive algorithm to calculate
\begin{enumerate}
\item the (restricted) reduced genus one Gromov--Witten invariants of $Y$;
\item the reduced genus one relative Gromov--Witten invariants of $(X,Y)$; 
\item the (restrirccted) reduced genus one rubber invariants of $\mathbb{P}=\PP_Y(\operatorname{N}_{Y|X} \oplus \OO_Y)$
\end{enumerate}
from the full genus zero and reduced genus one invariants of the ambient variety $X$.
\end{customthm}

This reconstruction result is most directly a generalization of Gathmann's calculation~\cite{Ga}. A higher genus reconstruction result for the ordinary theory is proved in a paper of Maulik and Pandharipande~\cite{MaulikPandharipande}, but is of a fundamentally different approach than ours. In particular, localization is never used in the analysis here, nor in the work of Gathmann.

We pursue a strategy laid out by Vakil, building on work of Caporaso and Harris~\cite{CH98,Vre}. The first step is to express the locus of maps where the contact order degenerates in terms of Chern classes of tautological vector bundles. The second step is to describe that locus in terms of fibre products of moduli spaces with smaller invariants.

Novelties are introduced in both steps. In the first, we describe the locus of degenerate maps as the zero locus of a logarithmic line bundle, arising from a piecewise linear function with a simple moduli theoretic description. The use of tropical techniques substantially simplifies Gathmann's calculations

We achieve the first step using techniques that are entirely within tropical and logarithmic geometry, substantially simplifying Gathmann's calculation. For the second step, we contend with the interaction of the relative splitting formula and the geometry of elliptic singularities, which is entirely new (\S \ref{subsection C0 splitting}). We express the factorisation condition -- central to the constructions in~\cite{RSPW,RSPW2} -- in terms of tautological classes, leading to Theorem~\ref{thm: recursion}.

\subsection{Broader contributions} Beyond genus one, we pass through some useful general techniques. First, we explain how to describe degenerate moduli spaces as fibre products. This is substantially more delicate than the standard situation for two reasons. On the one hand, we need to study the interaction of the relative geometry with elliptic singularities, which involves a study of the factorization condition in terms of tautological vector bundles. On the other hand, the degenerate strata of moduli spaces are not simply fiber products of spaces with smaller numerical invariants, but compactified torus bundles over these strata, due to the introduction of logarithmic blowups. This geometry is present throughout logarithmic enumerative geometry, particularly in the degeneration formula, and we provide a first systematic procedure that explains how to use tropical geometry to calculate with these torus bundles~\cite{R19}. In addition, the results of~\cite{RSPW} suggest the existence of a reduced higher genus Gromov--Witten theory formed by replacing contracted elliptic components with singularities\footnote{We have been informed by Jonathan Wise that his Ph.D. candidate Sebastian Bozlee has constructed these higher genus reduced Gromov--Witten invariants. A manuscript is in preparation.}. The discussion here is likely to carry over to this setting.

Second, a key step in our recursion is a description of the locus of maps with higher than prescribed tangency, which was identified by Gathmann. We realize the locus as the vanishing locus of a section of a line bundle that comes from tropical geometry — from a piecewise linear function on the fan. The systematic understanding of logarithmic line bundles arising in this fashion is likely to play an important role in logarithmic enumerative geometry.

We rephrase Gathmann's idea of increasing the tangency requirement within the framework of logarithmic stable maps (where maximal tangency is always assumed) by introducing \emph{fictitious markings}, and identifying the space of maps with higher contact order with a boundary stratum inside that of lower contact order and more (fictitious) markings.

@@

Finally, we note that our methods require us to encounter several variants of the space of relative/logarithmic stable maps, with both fixed and expanded target. The conceptual features of one space are often computational bugs, and vice versa, and we believe that the techniques developed in the genus one case here will be used repeatedly in logarithmic Gromov--Witten theory calculations. In particular, we frequently use Kato's perspective on logarithmic blowups as subfunctors that was applied in~\cite{RSPW,RSPW2}.

\subsection*{Acknowledgements} We have benefited from fruitful conversations with friends and colleagues, including Pierrick Bousseau, Francesca Carocci, Cristina Manolache, Davesh Maulik, Keli Santos-Parker, and Jonathan Wise. Important aspects of this work were completed while L.B. and N.N. were visiting MIT in Spring 2018 and the University of Cambridge in Winter 2019, and we thank these institutions for their hospitality. L.B. and N.N. would like to thank Imperial College London, the LSGNT, the Royal Society, 3CinG and the University of Glasgow for support. L.B. would in addition like to thank the MSRI program ''Enumerative Geometry beyond Numbers'', and MPIM-Bonn.

@@

\noindent
\textbf{Step 3.} Given a tropical map $F: \plC\to \mathbb R_{\geq 0}$ parameterized by $\Sigma^{\mathrm{Kim}}$, there is a largest radius $\delta$ (possibly equal to $0$) such that every vertex strictly contained in the circle of radius $\delta$ around the core has degree-marking $d=0$. Let $
\Sigma^{\mathrm{cen}}$ be the subdivision obtained by requiring that the vertices contained within the circle of radius $\delta$ around the core of $\plC$ are totally ordered. This involves introducing cones along which certain $\Z$-linear combinations of edge lengths are identified.\\
